Optimizing Abstract$^{\text 2}$ Machines

### 2012-11-01 paper/paper.tex

\documentclass{llncs}
\usepackage{alltt}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{xltxtra}
\setmonofont[Scale=MatchLowercase]{DejaVu Sans Mono}
%
\input{preamble}
\begin{document}
\title{Optimizing Abstract$^2$ Machines}
%
\frontmatter          % for the preliminaries
%
\author{J.~Ian Johnson\inst{1} \and Matthew Might\inst{2} \and David Van Horn\inst{1}}
\institute{Northeastern University \and University of Utah}

\maketitle              % typeset the title of the contribution

\begin{abstract}
The abstracting abstract machines approach to program analysis makes
it easy to design and implement sound and computable abstract
interpretations of higher-order programs.
%
However, the straightforward implementation is generally
not very fast.
%
This article contributes a step by step process for going from a naive
abstract$^2$ machine to an efficient program analyzer that is (nearly)
three orders of magnitude faster than the starting point, yet just as
precise (and purely functional).
\end{abstract}
%
\section{Introduction}

@@

\begin{figure}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\epsilon}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\\[2mm]
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mval,\msto}
\text{ where }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mlit,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\msto,\mvar,\mval)
\\
\ap{\mop,\mval,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\text{ where }\mkont\in\msto(\maddr)
\text{ and } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\caption{Abstract$^2$ machine for ISWIM}
\label{fig:aam}
\end{figure}

\paragraph{Concrete interpretation} can be characterized by setting the implicit
parameters of the relation given in Figure~\ref{fig:aam} as follows:
\begin{align*}
\mathit{push}(l,\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
\mbox{ where }\maddr \notin\msto
\\
\mathit{bind}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
\end{align*}
The resulting relation is non-deterministic in its choice of
addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming, this relation defines
a deterministic machine.  (The relation is really a function.)

@@

@@

\begin{figure}
\begin{alltt}
  ;; State \(\rightarrow\) Setof State
  (define (step state)
    (match state
      [(ev σ e ρ k)
       (match e
         [(var l x)           (for/set ((v (lookup ρ σ x))) (co σ k v))]
         [(num l n)           (set (co σ k n))]
         [(bln l b)           (set (co σ k b))]
         [(lam l x e)         (set (co σ k (clos l x e ρ)))]
         [(rec f (lam l x e)) (set (co σ k (rlos l f x e ρ)))]
         [(app l f e)
          (define-values (σ* a) (push state))
          (set (ev σ* f ρ (ar e ρ a)))]
         [(ife l e0 e1 e2)
          (define-values (σ* a) (push state))
          (set (ev σ* e0 ρ (ifk e1 e2 ρ a)))]
         [(1op l o e)
          (define-values (σ* a) (push state))
          (set (ev σ* e ρ (1opk o a)))])]
      [(co σ k v)
       (match k
         ['mt (set (ans σ v))]
         [(ar e ρ l) (set (ev σ e ρ (fn v l)))]
         [(fn f l)   (for/set ((k (get-cont σ l))) (ap σ f v k))]
         [(fi c a ρ l)
          (for/set ((k (get-cont σ l)))
            (ev σ (if v c a) ρ k))]
         [(oa o l)
          (for*/set ((k (get-cont σ l))
                     (v (delta o (list v))))
            (co σ k v))]
      [(ap σ fun a k)
       (match fun
         [(clos l x e ρ)
          (define-values (ρ* σ*) (bind state))
          (set (ev σ* e ρ* k))]
         [_ (set)])]
\end{alltt}
\caption{Implementation of machine transition relation.}
\end{figure}

@@

@@

\subsection{Generic fixpoint calculator}

@@

\subsection{Fixed-point specialization}

%% Should be commutted to beginning

@@

@@

\bibliographystyle{plain}
\bibliography{dvh-bibliography}

>>>>

\documentclass[preprint]{sigplanconf}
\usepackage{alltt}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}
% \usepackage{xltxtra}
% \setmonofont[Scale=MatchLowercase]{DejaVu Sans Mono}
%
\input{preamble}

\begin{document}

\conferenceinfo{WXYZ '05}{date, City.} 
\copyrightyear{2005} 
\copyrightdata{[to be supplied]}

% \titlebanner{banner above paper title}        % These are ignored unless
% \preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Optimizing Abstract$^{\text 2}$ Machines}

%% \authorinfo{J. Ian Johnson}
%%            {Northeastern University}
%%            {ianj@ccs.neu.edu}
%% \authorinfo{Matthew Might}
%%            {University of Utah}
%%            {might@cs.utah.edu}
%% \authorinfo{David Van Horn}
%%            {Northeastern University}
%%            {dvanhorn@ccs.neu.edu}
\authorinfo{}{}{}

\maketitle

\begin{abstract}
Recently \emph{abstracting abstract machines} has been proposed as a
systematic approach to designing sound and computable program analyses
for higher-order programming languages.  The approach has been applied
to a wide variety of languages with features widely considered
difficult to analyze

The abstracting abstract machines approach to program analysis makes
it easy to design and implement sound and computable abstract
interpretations of higher-order programs.
%
However, the straightforward implementation is generally
not very fast.
%
This article contributes a step by step process for going from a naive
abstract$^2$ machine to an efficient program analyzer that is (nearly)
three orders of magnitude faster than the starting point, yet just as
precise (and purely functional).
\end{abstract}

%% \category{CR-number}{subcategory}{third-level}

%% \terms
%% term1, term2

%% \keywords
%% keyword1, keyword2

\section{Introduction}

@@

\begin{figure*}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\\[2mm]
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mval,\msto}
\text{ where }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mlit,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\msto,\mvar,\mval)
\\
\ap{\mop,\mval,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\text{ where }\mkont\in\msto(\maddr)
\text{ and } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\caption{Abstract$^2$ machine for ISWIM}
\label{fig:aam}
\end{figure*}

\paragraph{Concrete interpretation} can be characterized by setting the implicit
parameters of the relation given in Figure~\ref{fig:aam} as follows:
\begin{align*}
\mathit{push}(\ell,\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
\mbox{ where }\maddr \notin\msto
\\
\mathit{bind}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
\mbox{ where }\maddr \notin\msto
\end{align*}
The resulting relation is non-deterministic in its choice of
addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming, this relation defines
a deterministic machine.  (The relation is really a function.)

@@

\paragraph 0CFA

\begin{align*}
\mathit{push}(\ell,\msto,\mkont) &= \ell,\msto\sqcup[\ell\mapsto\{\mkont\}]
\\
\mathit{bind}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar\mapsto\{\mval\}]
\end{align*}

@@

\begin{figure}
\begin{alltt}
  ;; State \(\rightarrow\) Setof State
  (define (step state)
    (match state
      [(ev σ e ρ k)
       (match e
         [(var\(\superscript\ell\) x)           (for/set ((v (lookup ρ σ x))) (co σ k v))]
         [(lit\(\superscript\ell\) l)           (set (co σ k n))]
         [(lam\(\superscript\ell\) x e)         (set (co σ k (clos x e ρ)))]
         [(app\(\superscript\ell\) f e)
          (define-values (σ* a) (push state))
          (set (ev σ* f ρ (ar e ρ a)))]
         [(ife\(\superscript\ell\) e0 e1 e2)
          (define-values (σ* a) (push state))
          (set (ev σ* e0 ρ (ifk e1 e2 ρ a)))])]
      [(co σ k v)
       (match k
         ['mt (set (ans σ v))]
         [(ar\(\superscript\ell\) e ρ) (set (ev σ e ρ (fn v l)))]
         [(fn\(\superscript\ell\) f)   (for/set ((k (get-cont σ l))) (ap σ f v k))]
         [(fi\(\superscript\ell\) c a ρ)
          (for/set ((k (get-cont σ l)))
            (ev σ (if v c a) ρ k))])]
      [(ap σ fun a k)
       (match fun
         [(clos l x e ρ)
          (define-values (ρ* σ*) (bind state))
          (set (ev σ* e ρ* k))]
         [(? op? o)
          (for*/set ((k (get-cont σ l))
                     (v (\(\Delta\) o (list v))))
            (co σ k v))
         [_ (set)]))]))
\end{alltt}
\caption{Implementation of machine transition relation.}
\end{figure}

@@

\begin{alltt}
(define \church2 (\(\lambda\) (f) (\(\lambda\) (x) (f (f x)))))

(define pred
  (\(\lambda\) (n)
    (\(\lambda\) (rf)
      (\(\lambda\) (rx)
        (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
          (\(\lambda\) (i) rx))
         (\(\lambda\) (id) id))))))
\end{alltt}

@@

\newpage
\subsection{Generic fixpoint calculator}

\begin{alltt}
;; Expr \(\rightarrow\) Setof State
(define (eval e)
  (fix step (set (inj e))))
\end{alltt}

@@

\newpage
\subsection{Fixed-point specialization}

\begin{alltt}
;; State^ -> State^
;; Specialized from wide-step : State^ -> { State^ } ≈ State^ -> State^
(define (wide-step-specialized state)
  (match state
    [(cons σ cs)
     (define-values (cs* σ*)
       (for/fold ([cs* (set)] [σ* σ])
         ([c cs])
         (match (step-compiled^ (cons σ c))
           [(cons σ** cs**)
            (values (set-union cs* cs**) (join-store σ* σ**))])))
     (cons σ* (set-union cs cs*))]))
\end{alltt}
%% Should be commutted to beginning

@@

\begin{alltt}
;; State^ -> State^
;; Specialized from wide-step : State^ -> { State^ } ≈ State^ -> State^
(define (wide-step-specialized state)
  (match state
    [(cons σ cs)
     (define-values (cs* ∆)
       (for/fold ([cs* (set)] [∆* '()])
         ([c cs])
         (match (step-compiled^ (cons σ c))
           [(cons ∆** cs**)
            (values (set-union cs* cs**) (append ∆** ∆*))])))
     (cons (update ∆ σ) (set-union cs cs*))]))
\end{alltt}

\[
\{ \widehat{v_3}, \widehat{v_5}, \widehat{v_6} \} = 0001011
\]

\[
\sigma = \mbox{\tt \#(0$_0$,\dots,0$_{|P|}$)}
\]

@@

\bibliographystyle{splncs}
\bibliography{dvh-bibliography}
